# Data {#sec-data}

```{r, include = FALSE}
source("R/booktem_setup.R")
source("R/my_setup.R")
```

Yes - this really is that important to deserve its own chapter.

## Getting Data

what are some relevant data-sets. \[ \] work in progress

Step 0: Save yourself a headache and collect your data in a processable format [https://open.substack.com/pub/statsepi/p/simple-tips-for-recording-data-in](https://open.substack.com/pub/statsepi/p/simple-tips-for-recording-data-in?utm_campaign=post&utm_medium=web) 

**Step 1: Data Wrangling**

-   Each row is an observation (usually a patient)
-   Each column contains only 1 type of data (more below)
-   No free text (if you need to, categorize responses)

## Formatting

Clean tabular format etc.

Use excel like a boss, if you're going to: More excel data https://cghlewis.com/blog/excel_entry/

Flat files: Flat files: https://evidence.dev/blog/what-is-a-flat-file?utm_campaign=Data_Elixir&utm_source=Data_Elixir_526

## Data types

**Step 2: For each data element, consider the data type**

-   Binary (aka dichotomous scale): e.g. Yes or No, 0 or 1
-   Unordered Categorical (nominal scale): e.g. Utah, Colorado, Nevada, Idaho
-   Ordered Categorical (ordinal scale): e.g. Room air, nasal cannula, HFNC, intubated, ECMO, dead
-   Continuous (interval & ratio scales - differ by whether 0 is special): e.g. Temperature (Celsius or Kelvin, respectively)

|           |                 |             |                     |              |
|-----------|-----------------|-------------|---------------------|--------------|
|           | **dichotomous** | **nominal** | **ordinal**         | **interval** |
| a.ka.     | binary          | categorical | ordered categorical | continuous   |
| n         | X               | X           | X                   | X            |
| \%        | X               | X           | X                   | X            |
| min       |                 |             | X                   | X            |
| max       |                 |             | X                   | X            |
| range     |                 |             | X                   | X            |
| mode      | X               | X           | X                   | X            |
| mean      |                 |             |                     | X            |
| median    |                 |             | X                   | X            |
| IQR       |                 |             | X                   | X            |
| Std. dev. |                 |             |                     | X            |
| Std. err. |                 |             |                     | X            |

From: Stoddard GJ. Biostatistics and Epidemiology Using Stata: A Course Manual. Salt Lake City, UT: University of Utah School of Medicine.

TODO: not sure this stuff should live here vs elsewhere:

**Step 3: Visualize the distribution of each data-point** (detect outliers, data entry errors, etc.)

Darren's hypothetical code lives in a spreadsheet "darren_proj.xlsx":

Here is some code that loads the excel spreadsheet into R (we'll revisit)

```{r}
#| echo: false
library(readxl)
library(dplyr)

# Define the URL of the Excel file on GitHub
url <- "https://github.com/reblocke/fellow_stats/raw/main/darren_proj.xls"

# Download the file temporarily
temp_file <- tempfile(fileext = ".xls")
download.file(url, temp_file, mode = "wb")

# Read the Excel file, select columns, and arrange
darren_data_sheet <- read_excel(temp_file) 

# Clean up by removing the temporary file
unlink(temp_file)
```

It's already (mostly) clean.

Let's summarize it:

```{r}
summary(darren_data_sheet)
```

Hmmm.. what's wrong with this?

R need to be told that the binary variables are binary (and not characters)

```{r}
library(dplyr)

# Convert 'y'/'n' in the splenectomy column to TRUE/FALSE
darren_data_sheet <- darren_data_sheet %>%
  mutate(splenectomy = ifelse(splenectomy == "y", TRUE, FALSE))

# Assuming darren_data_sheet is your dataframe
darren_data_sheet <- darren_data_sheet %>%
  mutate(`got_cteph?` = ifelse(`got_cteph?` == 1, TRUE, FALSE))

```

Let's visualize each element:

```{r}
library(ggplot2)

# First, the binary ones

# Plot for splenectomy
ggplot(darren_data_sheet, aes(x = factor(splenectomy))) +
  geom_bar() +
  labs(title = "Distribution of Splenectomy", x = "Splenectomy", y = "Count")

# Plot for prox_v_dist
ggplot(darren_data_sheet, aes(x = factor(prox_v_dist))) +
  geom_bar() +
  labs(title = "Distribution of Proximal vs. Distal", x = "Proximal vs Distal", y = "Count")

# Plot for got_cteph?
ggplot(darren_data_sheet, aes(x = factor(`got_cteph?`))) +
  geom_bar() +
  labs(title = "Distribution of CTEPH Diagnosis", x = "Got CTEPH?", y = "Count")


```

The categorical one:

```{r}
# Bar chart for hosp
ggplot(darren_data_sheet, aes(x = factor(hosp))) +
  geom_bar(fill = "coral", color = "black") +
  labs(title = "Distribution of Hospital", x = "Hospital", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Adjust text angle for better readability if needed
```

and finally, the continuous one:

```{r}
# Histogram for qanadli
ggplot(darren_data_sheet, aes(x = qanadli)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  labs(title = "Histogram of Qanadli Scores", x = "Qanadli Score", y = "Frequency")
```
